Contracts for unions and intersections are hard for Nickel

### 2021-05-29 paper.tex
Nice shape secs 5.1 and 5.2

The existing work that previously researched and implemented
union and intersection
as dynamic checks have, in our eyes, different shortcomings
that makes them suboptimal solutions.
In this section, we present informally present those issues,
starting with the foundational paper by Keil and Thiemann
\cite{KeilThiemannUnionIntersection}, then following to a recent
work by Williams, Morris, and Wadler \cite{RootCauseOfBlame}, and
finally going over some similar contracts available in the Racket
programming language.

@@

This straightforward solutions makes composing the Keil and Thiemann
solution with other existing systems more difficult; in the words
of Williams, Morris, and Wadler (see Section~\ref{sec:will-morr-wadl}):
``(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
special cases can hinder composition.''\cite{RootCauseOfBlame}
\info{(Yann) Does
it also make it less efficient? If $(A \& B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}
And it also could harm performance, or at least complicate an implementation,
since some sharing mechanism should be used to avoid checking the same contract,
over the same value, multiple times.\unsure{Arnaud says: this whole
  section is too abstract give concrete code and explain what goes
  wrong. It's fine to give code in Nickel syntax}

Another difficult issue Keil and Thiemann were able to solve, is the
fact that flat contracts and intersection do not usually play nice
together, as shown on Section \ref{sec:flat-and-inter}.
They provide two different sets of rules for their calculus,
the first one is non deterministic, and its purpose is to provide
semantics that are simple to understand.
With these rules, the interaction between flat contracts and
intersections becomes much simpler, since they are able to check
both sides of an intersection on separate threads.
This allows them to be sure that contracts from one branch of the intersection
won't get blamed while executing the other branch.

As an example, consider the programs shown on Figure
\ref{fig:kt-non-deterministic}, if the Original program
gets executed, it will spawn two instances of said program,
one with the first branch of the intersection (the First Program),
and the other one with the second one (the Second Program),
and these will get joined if they reach an equal execution point.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
x @ A & B
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=First Program]
x @ A
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Second Program]
x @ B
\end{lstlisting}
\caption{Non deterministic intersection contract opening}
\label{fig:kt-non-deterministic}
\end{figure}

The second set of rules is deterministic, and it is intended
to be simple to implement.
However, this second set needs to perform operations
on the evaluation context, to decide whether or not to evaluate
a given contract.
\footnote{You can check more on this topic on Section 5.1 of
\cite{KeilThiemannUnionIntersection}}

We see two problems here, a minor one is the fact that the two
different semantics defined for the calculus are really far apart
one from the other, both on the strategies used for contract resolution,
as well as on the complexity levels.
\resolved{(Yann) Is it really a
    problem per se? It is common thing to do (a nice declarative system
    for proofs and an algorithmic one for implementation, proved equivalent).
    Maybe the point is that the algorithmic system is way more complex than the
declarative one
(Teo) reworded}.
A second, major one,
is the dependency added over the evaluation context, since, again,
it makes composing this, with other features desired on programming languages,
difficult, or even impossible
\resolved{(Yann) I think we need to substantiate this claim. If this
context dependency explains why CSE is invalid, maybe have a little example?
(Teo) I removed it, I don't think there's an example, mainly since
this context dependency is not part of the language. I do think it complicates
implementation of said feature}
\todo{(Yann) TODO: see if there's a low hanging complexity
    argument we can make (like it's quadratic in the number of open contracts or
whatever)}.

@@

Williams, Morris, and Wadler, take a different approach on flat contracts,
they are completely forbidden in their system, explicitly to not encounter
the previously mentioned issue.
This restriction goes against what is usually expected from dynamic checks:
validation of data, and not just inclusion on types, is an expected
feature of contracts.

On top of that, a fundamental problem arises on said paper,
symptom of the difficulty of establishing good enough, and yet
simple semantics for union and intersection dynamic checks.
One of their main results, is the fact that their calculus obeys
all of their expected monitoring properties for higher ordered, union
and, intersection contracts.
However, one of these properties states the following:

$$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$

Which means that a continuation ($K$) complies negatively with an
intersection of $A$ and $B$ if it complies negatively with at least one of these.

Altough this is true, it's a weaker statement of what we'd like to have,
since it rules out continuations like $(\lambda f.~((f~3)~+~4,~not~(f ~ true)))$,
where $(\square 3)~+~4 \in \llbracket Number \rightarrow Number \rrbracket^-$
but $(\square 3)~+~4 \not \in \llbracket Boolean \rightarrow Boolean \rrbracket^-$,
and dually for $not~(f ~ true)$.
Which would mean, according to their semantics, that
the continuation presented above would not
be in $\llbracket Number \rightarrow Number \cap Boolean \rightarrow Boolean \rrbracket^-$,
even though our intuition (and the
introduction to the paper) indicates that it should.
\todo{Simplify example}

% since it rules out continuations like $(\lambda f.~(K_1[f],~K_2[f])) \square$,
% where $K_1 \in \llbracket A \rrbracket^-$ but $K_1 \not \in \llbracket B \rrbracket^-$,
% and dually for $K_2$, which would mean the continuation presented above would not
% be in $\llbracket A \cap B \rrbracket^-$, even though our intuition (and the
% introduction to the paper) indicates that it should.

While this mistake does not invalidate their work (the system they present
handles these cases correctly), it shows that existing research is still not on
good shape to handle the growing usecases being developed for
union and intersection around dynamic languages.

@@

% K ::= Id | K ◦ □ N | K ◦ V □ | K ◦ □ @pA
\unsure{Is it worth mentioning how continuations are formed?
Should we go so techincal?}
\todo{(Yann) Maybe adding a concrete example of this (it can be the same term but with
    concrete types like Number and String and simple contexts like application
    to a dumb argument) is sufficient to make the point}

>>>>

The existing work that previously researched and implemented
union and intersection
as dynamic checks have, in our eyes, different shortcomings
that makes them suboptimal solutions.
In this section, we informally present those issues,
starting with the foundational paper by Keil and Thiemann
\cite{KeilThiemannUnionIntersection}, then following to a recent
work by Williams, Morris, and Wadler \cite{RootCauseOfBlame}, and
finally going over some similar contracts available in the Racket
programming language.

@@

This straightforward idea makes composing the Keil and Thiemann
solution with other existing systems more difficult; in the words
of Williams, Morris, and Wadler (see Section~\ref{sec:will-morr-wadl}):
``(\ldots) the monitoring semantics for contracts of intersection and union types given by Keil
and Thiemann are not uniform. (\ldots) If uniformity helps composition, then
special cases can hinder composition.''\cite{RootCauseOfBlame}.
\info{(Yann) Does
it also make it less efficient? If $(A \& B)$ fails because of $B$, their
development causes to recheck the contract $A$ that has been duplicated?}
And it also could harm performance, or at least complicate an implementation,
since some sharing mechanism should be used to avoid checking the same contract,
over the same value, multiple times, exemplified by contract \nickel{A} on
Figure \ref{fig:kt-contract-transf}, that gets duplicated.
\unsure{Arnaud says: this whole
section is too abstract give concrete code and explain what goes
wrong. It's fine to give code in Nickel syntax}

Another difficult issue Keil and Thiemann were able to solve, is the
fact that flat contracts and intersection do not usually play nice
together, as shown on Section \ref{sec:flat-and-inter}.
They provide two different sets of rules for their calculus,
the first one is non deterministic, and its purpose is to provide
semantics that are simple to understand.
With these rules, the interaction between flat contracts and
intersections becomes much simpler, since they are able to check
both sides of an intersection on separate threads.
This allows them to be sure that contracts from one branch of the intersection
won't get blamed while executing the other branch, and vice versa.

As an example, consider the programs shown on Figure
\ref{fig:kt-non-deterministic}, if the Original program
gets executed, it will spawn two instances of said program,
one with the first branch of the intersection (\nickel{3 @ Number}),
and the other one with the second branch (\nickel{3 @ Positive}),
and these will get joined once they reach an equal point, in this case
it will be \nickel{3}.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
3 @ Number & Positive
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=First Program]
3 @ Number
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Second Program]
3 @ Positive
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Joined Program]
3
\end{lstlisting}
\caption{Non deterministic intersection contract opening}
\label{fig:kt-non-deterministic}
\end{figure}

The second set of rules is deterministic, and it is intended
to be simpler to implement.
However, this second set needs to perform operations
on the evaluation context, to decide whether or not to evaluate
a given contract.
\footnote{You can check more on this topic on Section 5.1 of
\cite{KeilThiemannUnionIntersection}}
For instance, in the program presented on Figure \ref{fig:inter-flat-contracts},
when using the value \nickel{(fun x => x) @ (String -> String)}
inside the execution of contract \nickel{C}, and applying it to \nickel{0},
the contract runtime would check the context of execution and it would
deduce that contract \nickel{String -> String} does not apply in this case,
skipping it.

We see two problems here, a minor one is the fact that the two
different semantics defined for the calculus are really far apart
one from the other, both on the strategies used for contract resolution,
as well as on the complexity levels.
\resolved{(Yann) Is it really a
    problem per se? It is common thing to do (a nice declarative system
    for proofs and an algorithmic one for implementation, proved equivalent).
    Maybe the point is that the algorithmic system is way more complex than the
declarative one
(Teo) reworded}.
A second, major one,
is the dependency added over the evaluation context, since, again,
it makes composing this, with other features desired on programming languages,
difficult, or even impossible.
\resolved{(Yann) I think we need to substantiate this claim. If this
context dependency explains why CSE is invalid, maybe have a little example?
(Teo) I removed it, I don't think there's an example, mainly since
this context dependency is not part of the language. I do think it complicates
implementation of said feature}
\todo{(Yann) TODO: see if there's a low hanging complexity
    argument we can make (like it's quadratic in the number of open contracts or
whatever)}

@@

Williams, Morris, and Wadler, take a different approach on flat contracts,
they are completely forbidden in their system, explicitly to not encounter
the previously mentioned issue.
This restriction goes against what is usually expected from dynamic checks:
validation of data, and not just inclusion on types.

On top of that, a fundamental problem arises on said paper,
symptom of the difficulty of establishing good enough, and yet
simple semantics for union and intersection dynamic checks.
One of their main results, is the fact that their calculus obeys
all of their expected monitoring properties for higher ordered, union
and intersection contracts.
However, one of these properties states the following:

\begin{figure}[h]
$$ K \in \llbracket A \cap B \rrbracket^-~if~K \in \llbracket A \rrbracket^- \lor K \in \llbracket B \rrbracket^- $$
\caption{Sound monitoring property for intersection from Williams, Morris, and Wadler.}
\label{fig:wmw-semantics}
\end{figure}

Which means that a continuation $K$ complies negatively with an
intersection of $A$ and $B$ if it complies negatively with at least one of these.
Or, less formally, that if you have a program that is missing a value at a given point,
you can put a value wrapped with a contract stating $A \cap B$,
and expect not to get a negative violation of that contract,
if you can either put a value
wrapped with $A$, or a value wrapped with $B$.

Altough this is true, it's a weaker statement of what we'd like to have,
consider the two contexts presented on Figure \ref{fig:valid-contexts},
the first one is a valid context for a \nickel{Number -> Number} function,
that is, you could replace the $\square$ by some function wrapped in a
\nickel{Number -> Number} contract, and the context will not violate it.
Similarly, the second context from the same figure is a valid
\nickel{Boolean -> Boolean} context.

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=\nickel{Number -> Number} context]
(*@$\square$@*) 3
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=\nickel{Boolean -> Boolean} context]
(*@$\square$@*) true
\end{lstlisting}
\caption{Two different contexts in Nickel}
\label{fig:valid-contexts}
\end{figure}

Next, we can combine these two contexts into one, that essentially
has one single missing value, that will be used independently in both contexts
from Figure \ref{fig:valid-contexts}, as shown on Figure \ref{fig:invalid-context}.
This context is not a valid \nickel{Number -> Number} context, since it will apply
the value put on $\square$ (through \nickel{f}) to a non \nickel{Number} (\nickel{true}).
In a similar way, it is not a valid \nickel{Boolean -> Boolean} context either.
Therefore, according to the semantics from the Williams, Morris, and Wadler paper, as
shown on Figure \ref{fig:wmw-semantics}, it is not a valid
\nickel{(Number -> Number) & (Boolean -> Boolean)} context.

\begin{figure}[h]
\begin{lstlisting}[language=nickel]
let f = (*@$\square$@*) in
(f 3, f true) 
\end{lstlisting}
\caption{Invalid \nickel{(Number -> Number) & (Boolean -> Boolean)} context,
combining contexts from Figure \ref{fig:valid-contexts}}
\label{fig:invalid-context}
\end{figure}

However, both our intuition, as the most common use case of intersections
(that is, overloading), tells us that this should work.
While this problem does not invalidate their work (in fact, the system they present
handles these cases correctly), it shows that existing research is still not on
good shape to handle the growing usecases being developed for
unions and intersections around dynamic languages.

@@

% K ::= Id | K ◦ □ N | K ◦ V □ | K ◦ □ @pA
\resolved{(Yann) Maybe adding a concrete example of this (it can be the same term but with
    concrete types like Number and String and simple contexts like application
    to a dumb argument) is sufficient to make the point}
